---
title: "Kinda, I guess, Advanced R"
author: "Mikhail Popov (mpopov.com)"
date: "October 15, 2014"
output:
 ioslides_presentation:
  widescreen: true
runtime: shiny
---

## Overview

- forward-pipe operator `%>%` aka "then-do"
- underused `%in%` operator
- `Reduce` function
- `...` argument for functions
- `<<-` operator and scope
- writing custom binary operators

## Forward-pipe operator aka "then-do"

The `magrittr` package adds a new operator to the R language and enables you to write cleaner code.

Traditionally, you write code inside-to-outside:

```{r,eval=FALSE}
h(f(x),y)
```

`%>%` lets you write a left-to-right pipeline:

```{r,eval=FALSE}
x %>% f %>% h(y) # Implict: h(<input>,y)
```

## *%>%* Example

```{r,echo=FALSE}
library(magrittr) # Provides pipe-forwarding mechanisms
library(knitr)
```

Nested functions:

```{r,eval=FALSE}
coef(summary(lm(sqrt(dist)~speed,cars)))
```

"then-do" approach:

```{r, echo=FALSE}
inputPanel(
  tags$textarea(id="ex1", rows=3, cols=40, style="width:700px;font-size:24px;", "")
)
renderPrint({
  eval(parse(text=input$ex1))
})
```

---

```{r,fig.height=4,fig.align='center'}
cars %>%
	transform(dist=sqrt(dist)) %>%
	function(x) { # anonymous function
			x %>% plot(ylab=expression(sqrt(dist)))
			x %>% lm(dist~speed,.) %>% abline
		}
```

---

`match` returns a vector of the positions of (first) matches of its first argument in its second:

```{r}
match(c("Shocktober","October","Rocktober","october"),month.name)
```

The binary operator `%in%` is a more intuitive interface to `match`:

```{r, echo=FALSE}
inputPanel(
  tags$textarea(id="ex2", rows=3, cols=40, style="width:700px;font-size:24px;", 'c("Shocktober","October","Rocktober","october") %in% month.name')
)
renderPrint({
  eval(parse(text=input$ex2))
})
```

---

`Reduce` uses a binary function to successively combine the elements of a given vector and a possibly given initial value.

```{r}
A <- month.name[2:6] # February, March, April, May, June
B <- month.name[4:10] # April, May, June, July, August, September, October
C <- month.name[3*(1:4)] # March, June, September, December
D <- month.name[2*(1:6)] # February, April, June, August, October, December
```

```{r, echo=FALSE}
inputPanel(
  tags$textarea(id="ex3", rows=3, cols=40, style="width:700px;font-size:24px;", 'intersect(intersect(intersect(A,B),C),D)')
)
renderPrint({
  eval(parse(text=input$ex3))
})
```

## All Together Now (Background)

Patients' `visits` to a clinic, with up to 5 diagnoses established per visit. Suppose a diagnosis is coded 001--100 (fake ICD-9).

```{r,echo=FALSE,results='asis'}
n <- 1e3
p <- 5
visits <- matrix(sample(sprintf("%03i",1:100),n*p,repl=T),nrow=n,ncol=p,byrow=T)
nais <- sample(2:5,n,repl=T,prob=(2:p)/sum(2:p))
for ( i in 1:n ) visits[i,nais[i]:p] <- NA
visits <- as.data.frame(visits); names(visits) <- paste("dx",1:p,sep="")
visits$age <- floor(rnorm(n,55,5))
visits %>% tail %>% kable
```

## All Together Now

### Objective

We are interested in patients whose "primary" (first three) diagnoses are kidney-related (049--060).

### Solution

```{r}
dx.kidney <- sprintf("%03i",49:60) # "049" "050" ... "059" "060"
index <- Reduce("|", # 'or'
				list(visits$dx1 %in% dx.kidney,
					 visits$dx2 %in% dx.kidney,
				   	 visits$dx3 %in% dx.kidney)) %>% which
# visits$age[index] : age of patients with kidney-related primary dx
```

## Optional arguments using '...'

**Example**: `apply(X, MARGIN, FUN, ...)` where `...` are arguments that are passed on to the function `FUN`

```{r}
inventory <- function(ArkID,...) {
	args <- list(...) # required
	args %>% unlist %>% as.table %>% addmargins
}
```

```{r, echo=FALSE}
inputPanel(
  tags$textarea(id="ex4", rows=3, cols=40, style="width:700px;font-size:24px;", 'inventory(1)')
)
renderPrint({
  eval(parse(text=input$ex4))
})
```

## Scope and <<-

A variable can be defined in global or local scope. Variables created inside of a function are local to it and die when the function is done. The operators `<<-` and `->>` cause a search to made through parent environments for an existing definition of the variable being assigned, subverting scope.


```{r, echo=FALSE}
inputPanel(
	tags$textarea(id="ex5", rows=3, cols=40, style="width:700px;height:200px;font-size:18px;", "save.creatures <- function(...) {
	args <- list(...) # required
	# some code that actually saves the creatures
	# creatures.saved <- args %>% unlist %>% as.table %>% sum 
	print(sprintf('Local: %i',creatures.saved))
}
creatures.saved <- 0

save.creatures(Cats=2,Dogs=2,Penguins=2,Dinosaurs=0,Snuggleupagus=1)

print(sprintf('Global: %i',creatures.saved))")
)
renderPrint({
	eval(parse(text=input$ex5))
})
```

## Binary Operators

| Operator                            | Description            |
| ----------------------------------- | ---------------------- |
| \+\, \-\, \*\, \/\, \^              | basic arithmetic       |
| x \%\% y                            | modulus (x mod y)      |
| X \%\*\% Y                          | matrix multiplication  |
| \<\, \<\=\, \>\, \>\=\, \=\=\, \!\= | basic logical          |
| \!x\, x \| y\, x \& y               | NOT x, x OR y, x AND y |
| x \%in\% y                          | elements of x in y     |
| x \%>\% f                           | apply function f to x  |

## Writing Custom Binary Operator

```{r}
"%rowInMat%" <- function(x,y) {
result <- logical(nrow(y))
for ( i in 1:nrow(y) ) result[i] <- all(x == y[i,]); return(result)
}
```

```{r, echo=FALSE}
inputPanel(
	tags$textarea(id="ex6", rows=3, cols=40, style="width:700px;font-size:24px;", 'set.seed(0); x <- matrix(sample(0:1,30,repl=T),ncol=5)
x
# c(1,1,1,1,1) %rowInMat% x')
)
renderPrint({
	eval(parse(text=input$ex6))
})
```
