---
output: html_document
---

*This is a post version of the talk to be given for Pittsburgh useR meeting on Wednesday, Oct 15.*

## Overview

- forward-pipe operator `%>%` aka "then-do"
- underused `%in%` operator
- `Reduce` function
- `...` argument for functions
- `<<-` operator and scope
- writing custom binary operators

## Forward-pipe operator *%>%* aka "then-do"

The `magrittr` package adds a new operator to the R language and enables you to write cleaner code.

Traditionally, you write code inside-to-outside:

```{r,eval=FALSE}
h(f(x),y)
```

`%>%` lets you write a left-to-right pipeline:

```{r,eval=FALSE}
x %>% f %>% h(y) # Implict: h(<input>,y)
```

By default, the input is given to the function as the first argument. You can also specify the position of the input with `.`

```{r,eval=FALSE}
x %>% f %>% h(y,.)
```

## *%>%* Example

Nested functions are hard to read and can get hard to write:

```{r,eval=FALSE}
library(xtable) # Export tables to LaTeX or HTML
xtable(
	coef(
		summary(
			lm(sqrt(dist)~speed,cars)
			)
		)
	)
```

The "then-do" approach makes writing and reading easier:

```{r,eval=FALSE}
library(magrittr) # Provides pipe-forwarding mechanisms
cars %>% lm(sqrt(dist)~speed,.) %>% summary %>% coef %>% xtable
```

```{r,echo=FALSE}
library(magrittr)
```

```{r}
library(knitr) # for kable() which renders tables to markdown
cars %>% lm(sqrt(dist)~.,.) %>% summary %>% coef
```

```{r,fig.height=4,fig.align='center'}
cars %>%
	transform(dist=sqrt(dist)) %>%
	function(x) {
		x %>% plot(ylab=expression(sqrt(dist)))
		x %>% lm(dist~speed,.) %>% abline
		}
```

## %in% operator

`match` returns a vector of the positions of (first) matches of its first argument in its second:

```{r}
match(c("Shocktober","october","Rocktober","October","Inktober"),month.name)
```

As you can see, the 4th element (October) is the 10th element in the vector of month names. The binary operator `%in%` is a more intuitive interface to `match`:

```{r}
c("Shocktober","october","Rocktober","October","Inktober") %in% month.name
```

```{r}
c("Shocktober","october","Rocktober","October","Inktober") %in% month.name %>% which
```

## *Reduce* function

```{r}
A <- month.name[2:6] # February, March, April, May, June
B <- month.name[4:10] # April, May, June, July, August, September, October
C <- month.name[3*(1:4)] # March, June, September, December
D <- month.name[2*(1:6)] # February, April, June, August, October, December
```
```{r}
intersect(intersect(intersect(A,B),C),D)
# alt: intersect(A,B) %>% intersect(C) %>% intersect(D)
```
That looks like it can get super messy super fast if we start adding even more elements. Fortunately, `Reduce` uses a binary function to successively combine the elements of a given vector and a possibly given initial value.
```{r}
Reduce(intersect,list(A,B,C,D))
```

### Silly Application

*Royal Rumble* is a battle royal that begins with two participants in the ring, with the remaining participants introduced one by one (at a set time period). The winner is the last participant standing. Take a moment to read the pipe sequence and figure out what's going on.
```{r,eval=FALSE}
library(babynames) # devtools::install_github("hadley/babynames")
library(dplyr) # A fast, consistent tool for working with data frame like objects.
top10 <- babynames %>% filter(year==1990) %>% top_n(10,n) %>% select(name) %>% sample
```
```{r,message=FALSE,warning=FALSE,echo=FALSE}
library(babynames)
library(dplyr)
set.seed(0)
top10 <- babynames %>% filter(year==1990) %>% top_n(10,n) %>% select(name) %>% c %>% unlist %>% sample
names(top10) <- paste("Contestant",1:10)
print(top10)
```
**Answer**: first we take the table of names and subset by the year 1990, then we get the top 10 names with respect to `n` (number of births), then we only get the column that's of interest to us, and finally we randomize the order. (The presented code is actually missing `%>% c %>% unlist %>%` between `select(name)` and `sample`.)
```{r,eval=FALSE}
fight <- function(x,y) sample(c(x,y),1,prob=c(2,1)/3)
Reduce(fight,top10)
```
```{r,echo=FALSE}
fight <- function(x,y) sample(c(x,y),1,prob=c(2,1)/3)
set.seed(78); top10 %>% Reduce(fight,.)
```

## Application of %in% and *Reduce* (Background)

Suppose we have patients' visits to a clinic, with up to 5 diagnoses established per visit. Suppose a diagnosis is coded 001--100. (Although in reality they would be ICD-9 codes.)

```{r,echo=FALSE,results='asis'}
n <- 1e3
p <- 5
visits <- matrix(sample(sprintf("%03i",1:100),n*p,repl=T),nrow=n,ncol=p,byrow=T)
nais <- sample(2:5,n,repl=T,prob=(2:p)/sum(2:p))
for ( i in 1:n ) visits[i,nais[i]:p] <- NA
visits <- as.data.frame(visits); names(visits) <- paste("dx",1:p,sep="")
visits$age <- floor(rnorm(n,55,5))
visits %>% tail %>% kable
```

## Application of %in% and *Reduce*

### Objective

We are interested in patients whose "primary" (first three) diagnoses are kidney-related (049--060).

### Solution

```{r}
dx.kidney <- sprintf("%03i",49:60) # "049" "050" ... "059" "060"
index <- Reduce("|", # 'or'
				list(visits$dx1 %in% dx.kidney,
					 visits$dx2 %in% dx.kidney,
					 visits$dx3 %in% dx.kidney)) %>% which
# visits$age[index] : age of patients with kidney-related primary dx
```

## Optional arguments using '...'

**Example**: `apply(X, MARGIN, FUN, ...)` where `...` are arguments that are passed on to the function `FUN`

```{r}
inventory <- function(ArkID,...) {
	args <- list(...) # required
	args %>% unlist %>% as.table %>% addmargins
	}
inventory(1,Noah=1,Cats=2,Dogs=2,Penguins=2,Dinosaurs=0)
```

## Out-of-scope assignment

A variable can be defined in global or local scope. Variables created inside of a function are local to it and die when the function is done. The operators `<<-` and `->>` cause a search to made through parent environments for an existing definition of the variable being assigned, subverting scope.

```{r}
save.creatures <- function(...) {
	args <- list(...) # required
	# some code that actually saves the creatures
	creatures.saved.local <- args %>% unlist %>% as.table %>% sum 
	creatures.saved.global <<- args %>% unlist %>% as.table %>% sum
	}
creatures.saved.local <- 0; creatures.saved.global <- 0
save.creatures(Cats=2,Dogs=2,Penguins=2,Dinosaurs=0,Snuggleupagus=1)
c(Local=creatures.saved.local,Global=creatures.saved.global)
```

## Binary Operators (built-in and in magrittr)

| Operator                            | Description            |
| ----------------------------------- | ---------------------- |
| \+\, \-\, \*\, \/\, \^              | basic arithmetic       |
| x \%\% y                            | modulus (x mod y)      |
| X \%\*\% Y                          | matrix multiplication  |
| \<\, \<\=\, \>\, \>\=\, \=\=\, \!\= | basic logical          |
| \!x\, x \| y\, x \& y               | NOT x, x OR y, x AND y |
| x \%in\% y                          | elements of x in y     |
| x \%>\% f                           | apply function f to x  |

### Writing Custom Binary Operator

```{r}
"%rowInMat%" <- function(x,y) {
	result <- logical(nrow(y))
	for ( i in 1:nrow(y) ) result[i] <- all(x == y[i,]); return(result)
	}
```

```{r,echo=FALSE}
set.seed(0); x <- matrix(sample(0:1,30,repl=T),ncol=5)
```

Suppose we have the following matrix `x`:
```
[,1] [,2] [,3] [,4] [,5]
[1,]    1    1    0    1    0
[2,]    0    1    1    0    0
[3,]    0    1    0    1    0
[4,]    1    1    1    1    0
[5,]    1    0    0    0    0
[6,]    0    0    1    1    1
```

```{r}
c(0,1,0,1,0) %rowInMat% x
```

